{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2023-03-25 Docs/manuscript.Rnw
added Sexpr in ms

\subsection{Analyses}
To estimate clines in spring and fall phenological events across species we used hierarchical Bayesian models. We regressed day of year (DOY) of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden. Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group. Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs. Europe) and species type (angiosperm vs. gymnosperm). All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}), using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples. We checked for model fit by ... [ADD here]. We present estimates as mean $\pm$ 90\% uncertainty intervals, unless otherwise stated. 
%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
%emw2Mar -- Did you use default priors? If so, can you figure out how to increase them threefold and see if answers change (they change run to run, but should not change dramatically)? Ideally they don't and then we can add 'increasing priors three-fold did not change results.'

@@

\todo[inline]{add map here}
\todo[inline]{add a table documenting studies attributes} %emw6Mar -- Tables should go in supplement. But we should add estimates to statements we make throughout. These should be added as mean +/- 90% uncertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically. 
Overall, spring events were not related to provenance latitude (insert the mean slope here, followed by the 90\% intervals) or mean annual temperature in North America (insert the mean slope here, followed by the 90\% intervals).  Spring events advanced slightly with provenance latitude and mean annual temperature in Europe (insert slope and UI), meaning spring events are slightly earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes)(Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).

@@

<<label=source_fall_models, echo=FALSE>>=
#load models and data to report numbers of things
#getwd()
source("../Analyses/source_fall_models_for_sweave.R")
@

In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower. This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(abs(median(northamericanpost_fitC_fall_lat)),digits = 2)} (UI = \Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.9)), digits = 2)}) days per degree we move north, or \Sexpr{format(median(northamericanpost_fitC_fall_mat),digits = 2)} days (UI) per degree decline in MAT$^{\circ}$C (Fig. \ref{figure:continent}). In Europe, such relationship is weak: advance \Sexpr{format(abs(median(europeanpost_fitC_fall_lat)),digits = 1)} (UI) days per degree we move north, or \Sexpr{format(median(europeanpost_fitC_fall_mat),digits = 1)} days (UI) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continent}).

@@

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms. However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}. Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(median(deciduouspost_fitC_fall_mat),digits = 2)} days VS. \Sexpr{format(median(evergreenspost_fitC_fall_mat),digits = 2)} days)(Fig. \ref{figure:spp_type}).

>>>>

\subsection{Analyses}
To estimate clines in spring and fall phenological events across species we used hierarchical Bayesian models. We regressed day of year (DOY) of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden. Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group. Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs. Europe) and species type (angiosperm vs. gymnosperm). All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}), using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples. We checked for model fit by ... [ADD here]. We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated. 
%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
%emw2Mar -- Did you use default priors? If so, can you figure out how to increase them threefold and see if answers change (they change run to run, but should not change dramatically)? Ideally they don't and then we can add 'increasing priors three-fold did not change results.'

@@

\todo[inline]{add map here}
\todo[inline]{add a table documenting studies attributes} %emw6Mar -- Tables should go in supplement. But we should add estimates to statements we make throughout. These should be added as mean +/- 90% uncertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically.

<<label=source_spring_models, echo=FALSE>>=
#load models and data to report numbers of things
# "C:/Users/alina/Documents/git/localadaptclim"
# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")
#getwd()
source("../Analyses/source_spring_models_for_sweave.R")
@

Overall, spring events were not related to provenance latitude (mean slope = \Sexpr{format(mean(northamerican_springpost_fitC_spring_lat),digits = 1)} $\pm$
\Sexpr{format(abs((quantile(northamerican_springpost_fitC_spring_lat, 0.95)-(quantile(northamerican_springpost_fitC_spring_lat, 0.05)))*0.5), digits = 2)}) or MAT (mean slope = \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} $\pm$
  \Sexpr{format(abs((quantile(northamerican_springpost_fitC_spring_mat, 0.95)-(quantile(northamerican_springpost_fitC_spring_mat, 0.05)))*0.5), digits = 2)}) in North America. \todo{add figure here}  Spring events advanced slightly with provenance latitude (mean slope = \Sexpr{format(mean(european_springpost_fitC_spring_lat),digits = 1)} $\pm$
  \Sexpr{format(abs((quantile(european_springpost_fitC_spring_lat, 0.95)-(quantile(european_springpost_fitC_spring_lat, 0.05)))*0.5), digits = 2)}) and MAT (mean slope = \Sexpr{format(mean(european_springpost_fitC_spring_mat),digits = 1)} $\pm$
  \Sexpr{format(abs((quantile(european_springpost_fitC_spring_mat, 0.95)-(quantile(european_springpost_fitC_spring_mat, 0.05)))*0.5), digits = 2)}) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).

@@

<<label=source_fall_models, echo=FALSE>>=
#load models and data to report numbers of things
# "C:/Users/alina/Documents/git/localadaptclim"
# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")
#getwd()
source("../Analyses/source_fall_models_for_sweave.R")
@

In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower. This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2)} (\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2)}-\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2)}) days per degree northward, or \Sexpr{format(mean(northamericanpost_fitC_fall_mat),digits = 2)} days (\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2)}-\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2)}) per degree decline in MAT$^{\circ}$C (Fig. \ref{figure:continent}). In Europe, such relationship is weak: advance \Sexpr{format(abs(mean(europeanpost_fitC_fall_lat)),digits = 1)} (\Sexpr{format(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 1)}-\Sexpr{format(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2)}) days per degree northward, or \Sexpr{format(mean(europeanpost_fitC_fall_mat),digits = 1)} days (\Sexpr{format(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2)}-\Sexpr{format(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2)}) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continent}).

@@

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms. However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}. Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(mean(deciduouspost_fitC_fall_mat),digits = 2)} days VS. \Sexpr{format(mean(evergreenspost_fitC_fall_mat),digits = 2)} days)(Fig. \ref{figure:spp_type}).

### 2023-03-25 Docs/manuscript.tex
added Sexpr in ms

\todo[inline]{add map here}
\todo[inline]{add a table documenting studies attributes} %emw6Mar -- Tables should go in supplement. But we should add estimates to statements we make throughout. These should be added as mean +/- 90% uncertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically. 
Overall, spring events were not related to provenance latitude (insert the mean slope here, followed by the 90\% intervals) or mean annual temperature in North America (insert the mean slope here, followed by the 90\% intervals).  Spring events advanced slightly with provenance latitude and mean annual temperature in Europe (insert slope and UI), meaning spring events are slightly earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes)(Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).

@@

In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower. This relationship, however, was observed mostly in North America where fall events advance 4.2 (UI = 4) days per degree we move north, or 6.4 days (UI) per degree decline in MAT$^{\circ}$C (Fig. \ref{figure:continent}). In Europe, such relationship is weak: advance 0.5 (UI) days per degree we move north, or 0.7 days (UI) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continent}).

>>>>

\todo[inline]{add map here}
\todo[inline]{add a table documenting studies attributes} %emw6Mar -- Tables should go in supplement. But we should add estimates to statements we make throughout. These should be added as mean +/- 90% uncertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically.

Overall, spring events were not related to provenance latitude (mean slope = 0.1 $\pm$
0.16) or MAT (mean slope = -0.09 $\pm$
  0.27) in North America. \todo{add figure here}  Spring events advanced slightly with provenance latitude (mean slope = 0.1 $\pm$
  0.33) and MAT (mean slope = -0.2 $\pm$
  0.39) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).

@@

In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower. This relationship, however, was observed mostly in North America where fall events advance 4.2 (3.9-4.5) days per degree northward, or 6.4 days (5.8-7) per degree decline in MAT$^{\circ}$C (Fig. \ref{figure:continent}). In Europe, such relationship is weak: advance 0.5 (0.2-1.2) days per degree northward, or 0.7 days (1-2.4) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continent}).
